---
title: "Analysis of Urban Tree Cover: City of Buffalo 2024"
author: Jack Mernitz
subtitle: Work in Progress- 'Pay no attention to the man behind the curtain'
date: today
date-format: long
format:
  html:
    theme: spacelab
    toc: true
    toc-location: left
    toc-title: "Content Sections"
    code-fold: true
---

# Introduction

Trees are often overlooked as a component of urban spaces. With so much focus on the growth of infrastructure, living space, and work spaces, green places are left little room to grow. Tree leaf area coverage, while possibly annoying for lawn owners in the fall, is directly related to shade capacity and other environmental benefits of trees. Permeable surfaces where trees grow also reduce runoff, trees cool the surrounding air in sufficient density, and the mental and aesthetic benefits of green spaces in urban areas for residents and tourists are notable. Urban forestry bureaus seek to classify and present some of these benefits to the public.

Quantifying and estimating the possible value of these services in relation to tree occurrence seemed like an interesting topic to delve into. Vacant spots and stumps of former trees cannot provide benefits any longer. While an intensive endeavor to replace fully grown trees, my project seeks to enumerate how useful it could be to increase tree cover. Parsing data to assign relative values for comparable trees will assist in more accurate estimation of services. Individual versus collective benefit will also be questioned, but only after calculation of total possible benefits are complete.

# Materials and methods

## The Dataset

Publicly available data from the City of Buffalo Department of Public Works: Bureau of Forestry exists as a list of tree inventory within the City proper. This includes multiple categories to represent trees' ecosystem services. Downloading and processing this data is necessary, as there are 133,229 entries in the dataset. The data is organized into 28 columns, and each row is a tree site. It is a daily updated dataset existing since 2018.

## Install and Load Packages

```{r, message=F, warning=F, results=F}
install.packages("mapview")
install.packages("hrbrthemes")
library(tidyverse)
library(dplyr)
library(ggplot2)
library(hrbrthemes)
library(mapview)
library(sf)
library(kableExtra)
library(htmlwidgets)
library(widgetframe)
knitr::opts_chunk$set(widgetframe_widgets_dir = 'widgets' ) 
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Downloading the required data

```{r}
tree_data <- read.csv("https://data.buffalony.gov/api/views/n4ni-uuec/rows.csv?accessType=DOWNLOAD")
```

### A sample of how the dataset is organized.

```{r, results='asis'}
tree_data %>% 
  slice(1:5) %>% #show only 1:5 rows
  kable(digits=2,align="c")%>% #make table and round to two digits
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive")) 
```

## Filter out 'unsuitable vacant' and '0' sites.
```{r}
tree_suitable <- tree_data %>%
  filter(tree_data[,2] != "unsuitable vacant")
tree_filtered <- tree_suitable %>%
  filter(tree_suitable[,3] != "0")
print(paste("The length of tree_filtered is ", nrow(tree_data) - nrow(tree_filtered), "rows smaller."))
```

### Resulting data split into vacant, stump, and used sites and compared.
```{r}
stump_data <- tree_filtered %>%
  filter(tree_filtered[,2] == "STUMP")

vacant_data <- tree_filtered %>%
  filter(tree_filtered[,2] == "VACANT")

treeonly_data <- tree_filtered %>%
  filter(!(tree_filtered[,2] %in% c("STUMP", "VACANT")))

stump_data %>%
  head(5) %>%
  kable() %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive"))

vacant_data %>%
  head(5) %>%
  kable() %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive"))

treeonly_data %>%
  head(5) %>%
  kable() %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive"))
```

### Species count analysis.

```{r fig.width=10, fig.height=6, warning=F}
treeonly_data %>%
  # Count occurrences of each tree species col3
  count(!!sym(names(treeonly_data)[3])) %>% 
  arrange(n) %>%  # Sort by count
  top_n(15, n) %>%
  mutate(!!names(treeonly_data)[3] := factor(!!sym(names(treeonly_data)[3]), levels = unique(!!sym(names(treeonly_data)[3])))) %>%  # Factor for ordered axis
  ggplot(aes(x = !!sym(names(treeonly_data)[3]), y = n)) +  # Use the count (n) on the y-axis
    geom_segment(aes(x = !!sym(names(treeonly_data)[3]), xend = !!sym(names(treeonly_data)[3]), y = 0, yend = n), color = "gray") +  # Add segments (bars)
    geom_point(size = 3, color = "#654194") +  # Add points (dots)
    coord_flip() +  # Flip coordinates to make the plot horizontal (lolipop graph)
    theme_ipsum() +  # Apply hrbrthemes style
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position = "none"  # Hide the legend
    ) +
    xlab("") 

top_15_counts <- treeonly_data %>%
  count(!!sym(names(treeonly_data)[3])) %>%
  arrange(desc(n)) %>%  # Sort by count in descending order
  slice_max(n, n = 15)  

#sum of 15 greatest
total_count <- sum(top_15_counts$n)

# Add the proportion column
top_15_counts <- top_15_counts %>%
  mutate(proportion = n / total_count)

portion_total <- total_count/70163
```
### Top 15 Species
```{r}
print(top_15_counts)
print(portion_total)
```
From this, we can also see the top 15 species only make up 57.8% of the total tree count.

### Assessment of resource values by species.
```{r}
top_eco <- treeonly_data %>%
  group_by(!!sym(names(treeonly_data)[3])) %>%  # Group by species
  summarise(total = sum(!!sym(names(treeonly_data)[5]), na.rm = TRUE)) %>%  
  arrange(desc(total)) %>%  # Sort by total value in descending order
  slice_max(total, n = 15)  # Keep top 15 categories by total sum

# Bar graph for top totals from column 5
top_eco %>%
  ggplot(aes(x = !!sym(names(top_eco)[1]), y = total, fill = as.factor(!!sym(names(top_eco)[1])))) +  # Use category for x-axis and fill by category
  geom_bar(stat = "identity") +  # Bar graph with stat = "identity" to use the actual totals
  geom_text(aes(label = total), angle = 90, hjust = .4, size = 5) +
  scale_y_continuous(limits = c(0, 1300000)) +
  theme(
    axis.text.x = element_blank(),   # Remove x-axis tick labels
    axis.ticks.x = element_blank()   # Remove x-axis ticks
  ) +
  labs(fill = "Common Name", title = "Ecological Yearly Benefits of Top 15 Species") +
  xlab("Trees") +
  ylab("Benefits in Dollars")
```
This shows the most ecologically beneficial tree species and their associated value. We'll also assume people do not want to go to the trouble of removing stumps first, so let's focus on vacant spots. We know that people will want to pick popular choices, so we'll use the 15 most popular species to assign trees to vacant sites.

# Results

## Proportionate species distribution assigned to vacant sites.
```{r}
# Calculate how many times each species should be assigned to vacant_data
vacant_count <- nrow(vacant_data)  # matches

# Calculate numb of times each species should appear in vacant_data
top_15_counts <- top_15_counts %>%
  mutate(count_vacant = round(proportion * vacant_count))

# Create a vector of species names, repeated according to count_vacant
reassigned_data <- rep(top_15_counts$Common.Name, top_15_counts$count_vacant)

# Randomly shuffle the species names in vacant_reassigned_data
set.seed(426)  # Set the seed to ensure reproducibility
reassigned_data <- sample(reassigned_data)

# Adjust the length of vacant_reassigned_data to match vacant_data
# Ensure it has exactly the same number of observations as vacant_data (60559)
reassigned_data <- sample(vacant_reassigned_data, vacant_count, replace = TRUE)

# Replace the "Common.Name" column (column 3) in vacant_data with reassigned species
vacant_data$Common.Name <- reassigned_data

vacant_data %>%
  head(5) %>%
  kable() %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive"))
```
We now have reassigned the 15 most populous trees to the vacant_data dataset.

## Take totals data and find eco value per tree.
```{r}
eco_summary <- treeonly_data %>%
  group_by(!!sym(names(treeonly_data)[3])) %>%  # Group by species (Common.Name)
  summarise(
    total_benefit = sum(!!sym(names(treeonly_data)[5]), na.rm = TRUE),  # Sum of eco benefit
    count = n(),  # Occurrences of each species
    avg_benefit = total_benefit / count  # Calculate the average ecological value
  ) %>%
  arrange(desc(count)) %>%  # Sort by count
  slice_max(count, n = 15)  # Top 15 species based on count
#Display
eco_summary
```

##Project environmental benefits of vacant site data replanted
```{r}
# Join the eco_summary with vacant_data based on the species (Common.Name)
vacant_eco <- vacant_data %>%
  left_join(eco_summary %>%
              select(!!sym(names(eco_summary)[1]), avg_benefit), # Select species and avg_benefit
            by = c("Common.Name" = names(eco_summary)[1])) %>%  # Join on species name
  mutate(!!sym(names(vacant_data)[5]) := avg_benefit)  # Replace the 5th column with avg_benefit

# Checked data, missing lat/long
vacant_eco <- vacant_eco[-12075, ]

# View the updated vacant data with the avg_benefit in the 5th column
vacant_eco %>%
  head(5) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```
## Total ecological benefit increase
```{r}
# Sum the total ecological benefit of vacant_eco
total_ecological_benefit <- sum(vacant_eco$avg_benefit, na.rm = TRUE)

# Print the total ecological benefit
print(total_ecological_benefit)

```
##Let's map the new trees!
```{r}
vacant_eco_sf <- vacant_eco %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)  # Convert to sf object

# Use mapview to visualize
mapview(vacant_eco_sf, zcol = "avg_benefit", legend = TRUE)
```


# Conclusions

Interesting analyses from count and total ecological benefits alone. Although higher in overall count, crabapple and planetree are not in the highest ecological benefit categories. That section is dominated by maple species. 

# References
